Supplementary for *lancer*: a R package for linearity assessment and visualisation of multiple curves

### 2023-01-22 Manuscript/manuscript.qmd
Update documents. See NEWS.md for more information.

Linearity assessment plays a significant role in the validation of instrumentation and experimental procedures. Linearity can be tested by applying several graphical and numerical approaches. Summary statistics of these curves are mostly limited to the Pearson Correlation Coefficient, which is insufficient to fully test the for linearity. In addition, spreadsheet software, e.g. Microsoft Excel, only allow the analyst to repetitively plot, view and analyse the linearity of curves one at a time, a tedious and time-consuming process. While plotting of many curves at once and calculation of additional summary statistics for assessing linearity of curves can be done using programming languages like R, implementing this from scratch can be arduous task for the analyst. As such, we created the R package *lancer* so that curves can be analysed efficiently with a few functions. To our knowledge, this is currently not implemented in other software or programming languages. *lancer* furthermore reports the statistical results in tables/spreadsheets and records the plots in a pdf file. In addition, *lancer* can also create interactive trellis plots, displayed as a HTML folder, for exploratory analysis, helping analysts when dealing with large datasets.

@@

Visual inspection of curves is useful but it must be accompanied by statistical tests as decision parameters. A commonly used test is the Pearson Correlation Coefficient. However, @sonnergaard2006 suggests that it is an ineffective standalone numeric parameter to estimate linearity. While researchers like @vanloco2002, @sanchez2021 and @logue2018 have indicated other metrics for linearity evaluation, these metrics are generally not implemented in most software. Furthermore, using spreadsheet software, e.g. Microsoft Excel, to individually plot numerous curves is time-consuming. While programming languages like R can plot numerous curve easily, it is hard to implement for a novice analyst.

The R package *lancer* addresses these issues by assisting analysts useful functions to plot many curves with additional metrics that better describe the curves' characteristics. It also provides analysts with an interactive viewer to group, filter and sort the curves, for examination of problematic cases, such as curves generated by saturated signals.

@@

Using a common column Sample Name, the two tables can be merged into one table (Curve Table) via `create_curve_table`.

Next, summary statistics are calculated via `summarise_curve_table` for each curve. Besides the Pearson Correlation Coefficient, an additional statistical test is the Mandel’s Fitting Test ($F_{stats}$ in @eq-mandel-test) as described by @andrade2013. The $F$ test can give evidence that a quadratic model fits better than a linear model, indicating a non-linear curve.

@@

Supplementary Figure 2 gives the summary statistics of three simulated curves (as examples), where one curve is linear and two curves show a plateau at high concentrations (denoted as saturation regime curve) and low concentrations (denoted as noise regime curve), respectively. The corresponding Pearson Correlation Coefficient values (`r_corr`) are $\ge0.9$ ($0.99$, $0.95$ and $0.98$ respectively) even for the non-linear curves. However, non-linear curves are detected by much lower Mandel’s Fitting Test $p$ values (`mandel_p_val`) ($1.66 * 10^{-4}$ and $2.56 * 10^{-3}$ respectively vs $0.38$ for the linear curve) and Percent Residual Accuracy values (`pra_linear`) ($62.30$ and $74.69$ respectively vs $94.32$ for the linear curve).

::: {#fig-overview layout="[[60,-40], [65,-35]]"}
![Curve Grouping Workflow](images/README-LinearEvaluation.png){#fig-linearity}

![Interactive Visualisation](images/README-TrellisOutput.png){#fig-visualisation}

*lancer*'s curve grouping workflow in @fig-linearity and interactive visualisation of curves in @fig-visualisation.
:::

After calculating the summary statistics for each curve, *lancer* uses the function `evaluate_linearity` to group the curves according to the workflows proposed in @fig-linearity. Workflow 1 uses the Pearson Correlation Coefficient and Percent Residual Accuracy to determine if the curve is linear (labelled as Good Linearity) or not (labelled as Poor Linearity). Workflow 2 goes one step further, using the Mandel’s Fitting Test and the fitted quadratic model’s concavity to check if a non-linear curve plateaus at low (labeled as Noise Regime) or high (labelled as Saturation Regime) concentrations. Non-linear curves that do not follow these trends are labelled as Poor Linearity.

@@

A benchmark workflow using only Pearson Correlation Coefficient value of $0.8$ is compared with Workflow 2 on simulated data sets of `r nrow(linear_curve_summary)` linear curves (labelled as Linear), curves that plateau at low (labelled as Noise Regime) or high (labelled as Saturation Regime) concentrations. Supplementary Figure 3 shows that Workflow 2 better identifies the saturation and noise regime curves than the benchmark workflow. While Workflow 2 correctly classifies a lower number of linear curves than the benchmark workflow, its percentage of correctly classified linear curves, `r inline_text(simulation_results, variable = wf2_group, level = "Good Linearity", column = "Good Linearity")`, is high. See <https://lancer-simulation.netlify.app> for report details. While the threshold values of Pearson Correlation Coefficient and Percent Residual Accuracy are based on the interpretation of @y.h.chan2003 and @logue2018, they remain subjective and arbitrary. Nevertheless, *lancer* allows optimization of these threshold values according to the analyst’s preference.

Although *lancer* can export the results in Excel or pdf, an interactive user interface can generate a better overview of the data. @fig-visualisation shows a HTML folder, exported by *lancer*, such that clicking on the index.html file inside the folder will open an interactive trellis plot that can group, filter and sort curves. This allows room for exploratory data analysis, such as identifying curves with linearity issues or understanding the effects of changing the Pearson Correlation Coefficient threshold to another value. Such information is hard to achieve with other common software. An example of an interactive viewer created by *lancer* can be found at <https://lancer-interactive-example.netlify.app>

@@

Linearity is one of the most important parameters of an analytical method to be evaluated. Our R package, *lancer*, can estimate linearity efficiently in high-throughput settings and with functions that plot many curves quickly, reporting curve summary statistics to better describe the shape of the datasets. It also provides an interactive trellis plot for exploratory data analysis. It is available on GitHub <https://github.com/SLINGhub/lancer> while the documentation and tutorials can be accessed from <https://slinghub.github.io/lancer>.

>>>>

Linearity assessment plays a significant role in the validation of instrumentation and experimental procedures. Linearity can be tested by applying several graphical and numerical approaches. Summary statistics of these curves are mostly limited to the Pearson Correlation Coefficient, which is insufficient to fully test the for linearity. In addition, spreadsheet software only allow the analyst to repetitively plot, view and analyse the linearity of curves one at a time, a tedious and time-consuming process. While plotting of many curves at once and calculation of additional summary statistics for assessing linearity of curves can be done using R, implementing this from scratch can be arduous task for the analyst. As such, we created the R package *lancer* so that curves can be analysed efficiently with a few functions. To our knowledge, this is currently not implemented in other software or programming languages. *lancer* furthermore reports the statistical results in tables/spreadsheets and records the plots in a pdf file. In addition, *lancer* can also create interactive trellis plots, displayed as a HTML folder, for exploratory analysis, helping analysts when dealing with large datasets.

@@

Visual inspection of curves is useful but must be accompanied by statistical tests as decision parameters. A commonly used test is the Pearson Correlation Coefficient. However, @sonnergaard2006 suggests that it is an ineffective standalone numeric parameter for accessing linearity. While @vanloco2002, @sanchez2021 and @logue2018 have indicated other metrics for linearity evaluation, these metrics are generally not implemented in most software. Furthermore, using spreadsheet software to individually plot numerous curves is time-consuming. While R (@rprogramming) can plot numerous curve easily, it is hard to implement for a novice analyst.

The R package *lancer* addresses these issues by assisting analysts with useful functions to plot many curves with additional metrics that better describe the curves' characteristics. @fig-visualisation shows that *lancer* also provides an interactive viewer to group, filter and sort the curves, for examination of problematic cases, such as curves generated by saturated signals.

![*lancer*'s interactive visualisation of curves](images/README-TrellisOutput.png){#fig-visualisation}

@@

Using a common column Sample Name, the two tables are merged into one table (Curve Table) via `create_curve_table`.

Next, each curve's summary statistics are calculated via `summarise_curve_table`. Besides the Pearson Correlation Coefficient, an additional statistical test is the Mandel’s Fitting Test ($F_{stats}$ in @eq-mandel-test) as described by @andrade2013. The $F$ test can give evidence that a quadratic model fits better than a linear model, indicating a non-linear curve.

@@

Supplementary Figure 2 gives the summary statistics of three simulated curves as examples, where one curve is linear and two non-linear curves with a plateau at high and low concentrations, denoted as saturation and noise regime curve respectively as defined by @galitzine2018. The corresponding Pearson Correlation Coefficient values (`r_corr`) are $\ge0.9$ ($0.99$, $0.95$ and $0.98$ respectively) even for the non-linear curves. However, non-linear curves are detected by much lower Mandel’s Fitting Test $p$ values (`mandel_p_val`) ($1.66 * 10^{-4}$ and $2.56 * 10^{-3}$ respectively vs $0.38$ for the linear curve) and Percent Residual Accuracy values (`pra_linear`) ($62.30$ and $74.69$ respectively vs $94.32$ for the linear curve).

After calculating the summary statistics for each curve, *lancer* uses the function `evaluate_linearity` to group the curves according to the workflows proposed in Supplementary Figure 3. Workflow 1 uses the Pearson Correlation Coefficient and Percent Residual Accuracy to determine if the curve is linear or not, labelled as Good or Poor Linearity respectively. Workflow 2 goes one step further, using Mandel’s Fitting Test and the fitted quadratic model’s concavity to label a non-linear curve as Saturation or Noise Regime, if it plateaus at high or low concentrations respectively. Non-linear curves that do not follow these trends are then labelled as Poor Linearity.

@@

A benchmark workflow using only Pearson Correlation Coefficient value of $0.8$ is compared with Workflow 2 on simulated data sets of `r nrow(linear_curve_summary)` linear curves (labelled as Linear), non-linear curves that plateau at high and low concentrations (labelled as Saturation and Noise Regime respectively). Supplementary Figure 4 shows that Workflow 2 better identifies the saturation and noise regime curves than the benchmark workflow. While Workflow 2 correctly classifies a lower number of linear curves than the benchmark workflow, its percentage of correctly classified linear curves, `r inline_text(simulation_results, variable = wf2_group, level = "Good Linearity", column = "Good Linearity")`, is high. See <https://lancer-simulation.netlify.app> for report details. While the threshold values of Pearson Correlation Coefficient and Percent Residual Accuracy are based on the interpretation of @y.h.chan2003 and @logue2018, they remain subjective and arbitrary. Nevertheless, *lancer* allows optimization of these threshold values according to the analyst’s preference.

Although *lancer* can export the results in Excel or pdf, an interactive interface can generate a better overview. @fig-visualisation shows a HTML folder, exported by *lancer*. Clicking on the `index.html` file inside the folder opens an interactive plot that can group, filter and sort curves. This allows room for exploratory data analysis, such as identifying curves with linearity issues or understanding the consequence of changing the Pearson Correlation Coefficient threshold to another value. Such information is hard to achieve with static plots alone. An interactive viewer example created by *lancer* can be found at <https://lancer-interactive-example.netlify.app>

@@

Linearity is one of the most important parameter of an analytical method to be evaluated. Our R package, *lancer*, can access linearity efficiently using functions that quickly plot and report curve summary statistics, which better describe the shape of the curve. It also provides an interactive plot for exploratory data analysis. *lancer* is available on GitHub <https://github.com/SLINGhub/lancer> while the documentation and tutorials can be accessed from <https://slinghub.github.io/lancer>.

### 2023-01-22 Manuscript/supplementary.qmd
Update documents. See NEWS.md for more information.

![*lancer* Workflow](images/README-Workflow.png){#fig-lancer-Workflow}

@@

@@

![Simulation Results](images/simulation_results.png){#fig-Simulation}

>>>>

![*lancer* Workflow](images/README-Workflow.png){#fig-lancer-workflow}

@@

![Curve Grouping Workflow](images/README-LinearEvaluation.png){#fig-linearity}

@@

![Simulation Results](images/simulation_results.png){#fig-simulation}
